---
title: "Lab 3: Visualizing 2D Data"
output: html_notebook
---
## 3.1 Coding technique: for-loops

When you want to repeat the same bit of code multiple times, a for-loop can be your friend. For example, if you would like to plot a T-S diagram for each station on one of the cruises, rather than type out the code for each station individually, we can just write it once and iterate or *loop* over the code. 

Let's start with a simple example:

### Example: print the statement "hello" four times
```{r}
for (ii in 1:4) {
  print("hello")
}
```

### Example: print the statement "hello" four times, including the number at the start of the statement e.g. "1 hello"
```{r}
for (ii in 1:4) {
  print(paste(ii,"hello"))
}
```

### Example: plot a T-S diagram for each station
*This is what Nick had last time. Probably needs changed here because we are reading the data in differently*

```{r}
for (i in 1:4) {
CTD <- read.csv(paste('20170912Colbysta',toString(i),'.csv',sep=""))
TSdiagram(CTD[,2],CTD[,3],xlim=c(31,33),ylim=c(5,15))
cat("press a key to continue")
blah <- readline()
}
```


*Other possible subsection: Line graphs for different stations (instead of bar charts)??*

# 4 Categorical data (t-tests, bar charts, box plots, other stats?)

## 4.1 t-tests
*a bit of a contrived example...*

1.	use discrete data and determine the mean chl conc for each station and depth
2.	use t-test to answer: is the surface chl conc different at station 1 compared to station 2

## 4.2 Bar charts

Bar charts for categorical data e.g. three different treatments and the photosynthetic rate at each treatment

## 4.3 Box plots
Box plots for categorical data where you have multiple measurements at each treatment so can do a mean, stdev, etc.






