---
title: "Lab 4: Visualizing 2D patterns in oceanographic cruise data: part 2"
output: html_notebook
---

This lab focuses on visualizing  temporal and spatial patterns in oceanographic cruise data with a focus on changes over time. The tools you are introduced to here can be used to graph and analyze data in your labs, your independent research projects and the cruise data gathered throughout the semester. Be sure to save any scripts that you develop so you can continue to use them.

# 1. Importing and wrangling data

Today, we are going to create plots of depth vs time for a given year and station, so we are going to need to wrangle our data into a different form compared to yesterday.

Let's start by importing the data as before.

```{r}
# read in the whole csv file
CTDdata <- read.csv('Colby_cruise_data_2012-2017.csv')
```

Our first big step is to format the dates into a more usable form.

```{r}
#making a list of the dates of each cruise
dates <- CTDdata[['Date']]

# making a list of the year
years <- floor(dates/10000)

# making a list of the months and days in mmdd format
monthdays <- dates - (years*10000)

# making a list of the months
mnths <- floor(monthdays/100)

# making a list of the days
days <- monthdays - (mnths*100)

# converting our dates to R date-times
Rdates <- ISOdate(years,mnths,days)

# determining the day of year (Julian Day) for each date
jday <- strftime(Rdates, format = '%j')

# add these to our dataframe
CTDdata$year <- years
CTDdata$month <- mnths
CTDdata$day <- days
CTDdata$doy <- jday
```

We just created a day-of-year variable which we are going to use as our x-axis. Depth will be our y-axis. We need to extract from the main `CTDdata` the data for a specific year and station before we can make our plot.

```{r}
# creating an index based on year
index <- (years == 2016)

# extracting our chosen year of data from CTDdata
yearData <- CTDdata[index,]

# extracting the data for our chosen data from yearData
stationData <- yearData[yearData$Station == 4,]

# making a list of the day of the year each cruise took place on
doy <- unique(stationData$doy)

#counting how many different days of year (i.e. cruises) we have
numCols <- length(doy)

#creating an empty matrix with 100 rows (i.e. for 100 depths) and a column for each cruise in this year
TEM <- matrix(NA,nrow = 100, ncol = numCols)

#looping through each cruise
for (i in seq_along(doy)){
  #looping through each depth interval (from 1m to 100m)
  for (j in 1:100){
    #extracting the relevant cruise from the stationData data frame
    cruiseData <- stationData[stationData$doy == doy[i],]
    
    #binned temperature value for given depth interval
    temp = mean(cruiseData[ceiling(cruiseData$Depth)==j,'Temperature']) 

    #assinging the binned temperature to the j-th row and st-th column of the TEM matrix
    TEM[j,i] <- temp
  }
}


```

# 2. Plotting temporal changes

We have manipulated our data into a form where we can now plot it. We are going to use the same plotting function as we used yesterday for doing the section plots.

```{r}
library("fields")

image.plot(x=as.numeric(doy),y=-100:-1,t(TEM[100:1,]),
           zlim=c(8,20),
           col=tim.colors(200),
           xlab='day of year',
           ylab='Depth (m)',
           main='Station 4, 2016',
           legend.lab='Temperature (C)',
legend.line=2.5)
```

Let's make this into a couple of funtions so it is easier to make similar plots for different stations or different years:

```{r}
#function for converting our data into a matrix for a given year and station (yesterday we did this for a given cruise)
col2matrix_yr_st <- function(CTDdata,yr,st,colName){
  #making an index based on year (provided by the yr argument)
  index <- (CTDdata$year == yr)
  
  #indexing the CTDdata using the above index
  yearData <- CTDdata[index,]
  
  #indexing the yearData based on the station (provided by the st argument)
  stationData <- yearData[yearData$Station == st,]
  
  #making a list of the day of years (for the x-axis)
  doy <- unique(stationData$doy)
  numCols <- length(doy)
  
  #creating an empty matrix with 100 rows (i.e. for 100 depths) and a column for each cruise in this year
  TEM <- matrix(NA,nrow = 100, ncol = numCols)
  
  #looping through each cruise
  for (i in seq_along(doy)){
    #looping through each depth interval (from 1m to 100m)
    for (j in 1:100){
      #extracting the relevant cruise from the stationData data frame
      cruiseData <- stationData[stationData$doy == doy[i],]
      
      #binned temperature value for given depth interval
      temp = mean(cruiseData[ceiling(cruiseData$Depth)==j,colName]) 
  
      #assinging the binned temperature to the j-th row and st-th column of the TEM matrix
      TEM[j,i] <- temp
    }
  }
  
  #we can only return one object from a function, but we want to return two... so in this case, we can put the two objects into a list and return the list
  returnList = list(mat = TEM,doy = doy)
  return(returnList)
}

#function to plot
sectionPlot <- function(x,z,zLims,title,colName){
  #DESCRIPTION OF ARGUMENTS:
  #x = x-variable, z = variable to plot in color,
  #zLims = limits of the colorscale,
  #title = title to give the figure
  #colName = label for the colorbar
  image.plot(x=x,y=-100:-1,t(z[100:1,]),
             zlim=zLims,
             col=tim.colors(200),
             xlab='day of year',
             ylab='Depth (m)',
             main=title,
             legend.lab=colName,
  legend.line=2.5)
}

```

Let's try out these functions for a different year:

```{r}

output <- col2matrix_yr_st(CTDdata,2017,4,'Temperature')
sectionPlot(x=as.numeric(output$doy),z = output$mat, zLims = c(8,20), 
            title = 'Station4, 2017',colName = 'Temperature (deg C)')

```

Or what about a plot for all the stations in 2016:
```{r}

for (i in 1:4){
  output <- col2matrix_yr_st(CTDdata,2016,i,'Temperature')
  sectionPlot(x=as.numeric(output$doy),z = output$mat, zLims = c(8,20), 
              title = paste('Station ',i,', 2016', sep=''),colName = 'Temperature (deg C)')
}
```




# Assignment

We will assign you each a discrete variable (from the bottle data) and one continous variable (from the CTD profiles) and a year. 

For your assigned variables & year:

1. Plot
    1. A contour plot (depth vs station) of the discrete variable *for each cruise*
    2. A section plot (depth vs station) of the continuous variable *for each cruise*
    3. A temporal change plot (depth vs day of year) of the discrete variable *for station 4*
    4. A temporal change plot (depth vs day of year) of the continuous variable *for station 4*
    
2. Write a couple of paragraphs of a results + discussion section of a paper. This should cover the following:
    1. How do the variables change with respect to space (vertically/horizontally) and time?
    2. How do the changes in the two variables relate to each other?
    3. What ocean processes underly these changes?


### Due Friday 28th at 10pm. Email your write-up and figures (cmitchell@bigelow.org)