---
title: "Lab 3: Visualizing XXXXXXXXXXXXx"
output: html_notebook
---

This lab focuses on visualizing XXXXXXXXXXXXXXX. The tools you are introduced to here can be used to graph and analyze data in your labs, your independent research projects and the cruise data gathered throughout the semester. Be sure to save any scripts that you develop so you can continue to use them.

# 1. Importing and wrangling data

Today, we are going to import and wrangle the CTD data in the same way as yesterday.

Let's start by importing the data as before.

```{r}
# read in the whole csv file
DATA <- read.csv('Colby_cruise_data_2012-2017.csv', header = TRUE)
```

Now, let's make the `cruiseList` like yesterday:


```{r}
#making a list of the dates of each cruise
dates <- unique(DATA[['Date']])

#creating an empty list which we will fill with more lists in a for-loop
cruiseList <- list()

#loop to extract each cruise from the main DATA data frame and to group the individual cruise lists into one master list: cruiseList
for (dd in dates){
  #extracting a specific cruise into a list from the main DATA data frame
  cruiseData <- DATA[DATA[['Date']] == dd,]
  
  #making a unique cruise id
  id <- paste("colby",dd,sep="")
  
  #making a list of all the stations on this cruise
  stations <- unique(cruiseData$Station)

  #creating an empty list to be filled with each station on the next for-loop
  stationList <- list()
  
  #loop to extract each station from the cruiseData dataframe and to group the individual stations into one list for a given cruise: stationList 
  for (ss in stations){
    #extracting a specific station from the cruiseData dataframe
    stationData <- cruiseData[cruiseData$Station == ss,]
    
    #making a station id
    stationID <- paste('station',ss,sep = '')
    
    #adding the stationData to the station list, naming it with the station id
    stationList[[stationID]] <- stationData
  }

  
  #adding the staton list to the master cruiseList, naming it with the unique id
  cruiseList[[id]] <- stationList
}
```

# 2. Plotting oceanographic sections: take 2

Yesterday, we plotted sections for a given cruise i.e. we plotted sections of depth vs latitude. What about a section for a given station over time?

#### Example: plot a temperature section for station 1 from all the available cruise data

```{r}
cruises <- names(cruiseList)

stationData <- list()

for (cr in cruises){
  stData <- cruiseList[[cr]][['station1']]
  stationData[[cr]] <- stData
  }

#stationDataFrame <- bind_rows(stationData)
```

```{r}
library("fields")
```

```{r}
dates <- unique(stationDataFrame$Date)
cruises <- names(stationData)

TEM <- matrix(NA,nrow = 100, ncol = length(cruises))

for (c in 1:length(cruises)){
  for (j in 1:100){
    stData <- stationData[[cruises[c]]]
    
    TEM[j,c] <- mean(stData[ceiling(stData$Depth)==j,'Temperature']) 
  }
}

par(las=2)
image.plot(x=1:length(cruises),y=-100:-1,t(TEM[100:1,1:length(cruises)]),
           zlim=c(8,20),
           col=tim.colors(24),
           xlab='',
           ylab='Depth (m)',
           main='station1 : Temperature (C)',
legend.line=2.5)
axis(1, at = 1:length(cruises),labels=cruises)
```



*Other possible subsection: Line graphs for different stations (instead of bar charts)??*

# 4 Categorical data (t-tests, bar charts, box plots, other stats?)

## 4.1 t-tests
*a bit of a contrived example...*

1.	use discrete data and determine the mean chl conc for each station and depth
2.	use t-test to answer: is the surface chl conc different at station 1 compared to station 2

## 4.2 Bar charts

Bar charts for categorical data e.g. three different treatments and the photosynthetic rate at each treatment

## 4.3 Box plots
Box plots for categorical data where you have multiple measurements at each treatment so can do a mean, stdev, etc.






